Stability and Performance of Coalitions of Prosumers Through Diversification in the Smart Grid

### 2015-08-03 Article.tex
cut the...
cut

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot  \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, michel.marot, monique.becker\}@telecom-sudparis.eu}}

@@

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}).

Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of risk acceptable. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents (a task often studied through game theory). Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the returns distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}.

Nevertheless, one of the key point in the Markowitz theory is to consider explicitly the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interesting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}.

In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following. A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper.

More formally, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

@@

Where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $.

For a prosumer i, it is possible to write both quantities as a sum over the distributed energy resources ($ DER_{i} $) and loads ($ load_{i} $) of i :

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar radiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix \ref{appendixB} and in \cite{Lydia2014}. The process for generating the $ P_{i} $ series is pictured in the first block of the process diagram (see figure \ref{fig:process}).

\begin{figure}
\includegraphics[scale=.49]{Fig2}
\caption{Process diagram}
\label{fig:process}
\end{figure}

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single wind-turbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) through more complex combinations. In practice, we use random configurations for the agents.

In the rest of the paper, we use french weather data \cite{Infoclimat} starting in January 2006 and ending in December 2012, with a sampling frequency of three hours, and generate N timeseries of extra-production over this date range.

@@

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected productions but presenting a high unpredictability are penalized and can only afford small contracts.

In order not to overload the market with unrealistically small coalitions, the grid operator also specifies a lower bound $ P^{MIN} $ on the contract values. We thus characterized a valid coalition as one satisfying the two conditions :

@@

\begin{equation}
gain(S) = \int_{t_{0}}^{t_{k}} \lambda P_{S}^{CRCT \star} dt = P_{S}^{CRCT \star} \int_{t_{0}}^{t_{k}} \lambda dt
\end{equation} 
(since $ P_{S}^{CRCT \star} $ is supposed to be a constant rate over the contracted period). Using $ gain(S) $ directly as a utility function suffers a major drawback. It is indeed not a concave function of the coalitions' sizes, meaning that coalitions can grow as large as the number of agents allows it, without any counterbalance effects.

Such a model, that virtually allows infinitely large coalitions and contract values, is in practice not realistic. There are indeed costs (communication costs for instance) that increase with the coalitions sizes. We take this observation into account by rescaling the utility of a coalition S by its size in term of number of agents ($|S|$):

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT \star} }{P^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}
where parameter $ \alpha $ controls to what extent the size of a coalition impacts its utility, and $ P^{MAX} $ is a normalizing factor. $ P^{MAX} $ can be seen as the maximum production which can be injected in the grid.

Based on $ \mathcal{U} $, the marginal contribution of an agent i can be expressed as $ \delta_{S}(i) = \mathcal{U}(S+\{i\}) - \mathcal{U}(S) $. A coalition S has thus an interest in adding an additional agent i if this marginal contribution is positive :

@@

If $ \alpha $ is set to zero, agents are added as long as they increase the contract value of the coalition. If $ \alpha $ is greater than zero, additional agents have to increase the contract value by some factor. The utility function with $ \alpha $ is not necessarily convenient, here we relate $ \alpha $ to the mean sizes of the coalitions $ \bar{N} $. otherwise $ \mathcal{U}(S) $ tends to form coalitions of size approximately in the order of $ \bar{N} $, then :

@@

Figure \ref{fig:mean_approx} shows how $ \alpha^{\star} $ and the utility function evolves according to the mean size of the coalitions $ \bar{N} $. These curves are only valid in the simplified Gaussian example considered here. Nonetheless, they will provide some guidance when using real data.

@@

As can be pointed out, the purpose of $ \mathcal{U} $ is not to study coalitions stability against player defection which could be done through game theory, nor to redistribute the coalition's utility in terms of individual payoffs. But we aim to design $ \mathcal{U} $ as a measure of how good a given coalition is according to our criteria. In other terms does a given coalitions has a good production to risk ratio.

@@

\subsection{Representing the correlation structure}
As seen in section \ref{sec:notations}, the variance of the aggregated production impacts directly the contract values, and depends on the covariances between the agents productions. We argue here that, by having some representation of the correlation structure between the agents, the search landscape for high utility coalitions could be reduced, such that good coalitions are more likely to be found quickly. Usually, this correlation structure is formalized with a covariance matrix or a correlation matrix that contains all the correlation coefficients between the agents : $ M = (\rho_{ij})_{\forall i,j \in \mathcal{A}^{2}}$. By using a metric to map this matrix in a weighted adjacency matrix (see section \ref{sec:related}), it is possible to obtain a graph representation of the correlation relationships between the agents.

In the following, we use two opposite distance metrics for this mapping :

@@

Clearly, $ d^{1} $ (resp. $ d^{2} $) maps two correlated series as close points (resp. distant) while two uncorrelated series are distant (resp. close). These metrics enable us to compute a correlation graph $ G_{1} = (\mathcal{A}, E_{1}) $ and a "de-correlation" graph $ G_{2} = (\mathcal{A}, E_{2} ) $. For any i and j, the weight of the edge $ e_{ij} $ is $ d_{ij}^{1} $ in $ G_{1} $ and $ d_{ij}^{2} $ in $ G_{2} $.

In both cases, we want to keep only the edges which weights are located in the lower tail of the distance distributions. In other words, we want to compute the $ \epsilon$-graphs of $ G_{1} $ and $ G_{2} $ such that only meaningful edges remain. Selecting the right filter $ \epsilon $ is thus an important point since it affects the landscape search for the coalition formation. Unfortunately, there seems to be no clear consensus in the literature on how to select such a threshold. We will see later in this section that cliques in $ G_{2} $ are potential seeds for the coalitions. Since we want to generate $ N_{COAL} $ coalitions, we need at least $ N_{COAL} $ cliques of a given size to start. Besides, since we consider coalitions as disjoint, the starting cliques should be non overlapping. We thus select our optimal threshold for $ G_{2} $ as :

@@

Nevertheless de-correlation seems like a more complex concept than correlation in the sense that there is not even a partial notion of transitivity when it comes to it. Therefore, the clustering coefficients of $ G_{1}^{\epsilon} $ is much higher than the one of $ G_{2}^{\epsilon} $. This can be seen as another formulation of \cite{Garas2008} on the structural roles of weak and strong links on financial correlation graphs. Strong links, accounting for strong correlation relationships, are responsible for the clustering, while weak links provide the connectivity between clusters. Searching for clusters in $ G_{2}^{\epsilon} $ and hoping that this strategy will provide a nice coalition structure of internally uncorrelated coalitions seems thus pointless.

Consider now a clique in  $ G_{2}^{\epsilon} $, which is a complete subgraph of $ G_{2}^{\epsilon} $. This is indeed a structure of interest for our purpose. Since there is a link for every pairs of nodes, we know, by construction, that a clique has a mean correlation and a maximum correlation less than $ \epsilon $.

\begin{figure}
\includegraphics[scale=.6]{cliques_histo}
\caption{Histograms of utility values for coalitions of size 3. Red bars stands for cliques in the decorrelation graph, and blue bars represents all the other possible triplets. As visible, cliques tend to exhibit higher utilities than randomly selected coalitions.}
\label{fig:histo_cliques}
\end{figure}

Figure \ref{fig:histo_cliques} shows the distributions of the utility values for cliques of size 3 (triangles) in $ G_{2}^{\epsilon^{\star}} $ and for all the other possible triplets of agents. It is clearly visible that cliques tend to exhibit higher utilities because of their de-correlation property. Choosing cliques in $ G_{2}^{\epsilon^{\star}} $ as coalitions seems therefore appealing. Nevertheless, the quality of the results seems to decrease as the sizes of the cliques increase. Indeed, the larger the desired cliques, the more dense $ G_{2}^{\epsilon^{\star}} $ becomes (see equation \ref{epsilon_star}). There is a point where cliques results more from noisy edges than true de-correlation, which decreases the quality of the results.

Directly mapping cliques to coalitions by this de-correlation oriented approach is thus not sufficient. It is indeed possible that adding agents to these cliques has the combined effect of increasing the expected production while decreasing its stability. The question revolves around measuring the benefits of this production surplus compared to the disadvantage of having coalition with high volatility. This can be quantified by the marginal benefit in equation \ref{eq:marginal_benefit}.

@@

\begin{itemize}
\item \textbf{Random sampling of coalitions :} Coalitions are formed randomly without any other constraint that the desired size. This enables us to have an idea about the distributions of utility values for coalitions of a given size.
\item \textbf{Random sampling of coalition structures :} Coalition structures are sampled randomly by shuffling and random divisions of the agents. Algorithm \ref{alg:algo2} uses such a sampling and returns the highest utility coalition structure sampled. This algorithm will be refered to as "\textit{random}" in the following.
\item \textbf{Correlated :} This is the complete opposite of our algorithm. It uses the correlation graph $ G_{1} $ and performs a community detection. The resulting coalitions have thus very high internal correlations. We thus expect this algorithm to perform very bad compared to the others. See algorithm \ref{alg:algo3}.
\end{itemize}

@@

\begin{figure}
\includegraphics[scale=.6]{real_utility_2}
\caption{Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to equation \ref{eq:alpha_star} in order to favor 40 agents coalitions.}
\label{fig:real_utility2}
\end{figure}

\begin{figure}
\includegraphics[scale=.6]{search}
\caption{Evolution of the global utility $ \mathcal{U}(CS) $ (blue diamond curve, left axis) and the number of agents involved in the coalitions (red circle curve, right axis) during the greedy optimization of algorithm \ref{alg:algo1} }
\label{fig:search}
\end{figure}

Figure \ref{fig:coalitions} shows the coalitions formed with the considered algorithms in the contract value / volatility space. The color map in the background indicates regions where we expect high utilities (red) and the ones where we expect very poor utility values (blue). The bottom right corner, with high contract values and low volatilities, is therefore the region where we wish to form our coalitions. A single coalition is represented by a marker and the color and shape of a marker indicates by which algorithm the coalition has been formed. Besides, the sizes of the coalitions are indicated on the markers, and the marker size is also proportional to the coalition size. We can see that the utility function results in approximatively balanced coalitions. Small yellow markers indicates the gravity centers of their respective coalition structures. The coalitions of correlated agents (green squares) are clearly of poor quality according to our criteria since they can only afford small production contracts, and with a very high volatility.

On figure \ref{fig:coalitions}, the decorrelated coalitions (blue dots) are closer to the bottom right corner indicating a much better quality in term of productivity over volatility ratio. The black dotted line indicates the mean values for the random coalitions sampling technique. Each small dot stands for the mean position of all sampled coalitions of this given size. Variances are not indicated for readability, but are usually quite large since this sampling only takes the size as a constraint. We can see that as coalitions get larger, they tend to increase on average their contract values, but at the price of a higher volatility.

On figure \ref{fig:coalitions}, the results of the random coalition structure sampling are shown with the red ellipses that represent the distribution of the gravity centers of the sampled structures. Since the center of the ellipses stands for the mean and each ellipse adds one standard deviation, more than $ 99\% $ of the sampled gravity centers are within the largest ellipse. The small yellow dot below the ellipses indicates the gravity center of our solution. It is thus visible that our greedy graph based algorithm is able to find a quite good coalition structure in terms of volatility and contract values.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot}

@@

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}). Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of risk acceptable. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents (a task often studied through game theory). Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

These are only a few possibilities for balancing production and demand in the smart grid. Most of the time, these techniques will be coupled with predictions of the upcoming load curves and weather conditions. Combining all these technologies enables aggregators to quantify their expected production and the inherent risk that comes with it. The optimization of expected returns to risk is a traditional goal in finance, and a wide literature exists on this topic. It is well-known for instance, that the more risk one is willing to take, the higher his potential gains. On the contrary, when investing exclusively on low risks assets, one should expect relatively small gains. This trade-off is formalized in the Markowitz' portfolio theory \cite{Markowitz1959}. More precisely, given a set of assets for which we have some historic data of returns, the objective is to find a linear combination of these assets (the so-called portfolio) which maximizes the expected value while minimizing the variance of the portfolio's return. Markowitz's answer is a set of efficient portfolios that all optimize in some sense this trade-off. If one is able to put a number on his risk acceptance or on the target expected return, the corresponding efficient portfolio is a priori the best option. One of the most controversial assumptions in the portfolio theory is that returns are jointly normally distributed (or, at least, that the returns distribution is jointly elliptical). Some economist have pointed out the fact that this assumption might not capture well the reality of financial markets \cite{doi:10.1142/S0219024912500197}. Nevertheless, one of the key point in the Markowitz theory is to consider explicitly the correlation between the assets since they impact directly the variances of the portfolios. Since the work of \cite{Mantegna1999}, an interesting approach consists in computing a distance metric based on the correlation coefficients in order to organize the series in a correlation graph. Nodes represent the series considered while the edges are weighted by the metric. Because the metric can be computed for all pairs, these graphs are complete and of little use as is. Historically, the approach used by \cite{Mantegna1999} was to compute a minimum spanning tree as to obtain a hierarchical clustering of the series. Later on, it was pointed out that, by definition, a spanning tree could not capture the underlying clustering structure hidden in the correlation graph. In this paper, we use another classical filtering technique called $ \epsilon $-graph \cite{Garas2008}. It consists in selecting a threshold $ \epsilon $, and filtering out edges with smaller weights. As we will see further in this paper, this approach has the advantage of preserving clusters of correlated series.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following. A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

@@

Where $ P_{i}^{P}(t) $ represents the total production of agent i at time t and $ P_{i}^{D}(t) $ its consumption at time t. In other words, $ P_{i}(t) $ represents the instantaneous surplus of power that agent i is willing to sell at time t. As explained above, since large datasets containing this quantity over time are not yet available, we simulated these traces by considering separately $ P_{i}^{P} $ and $ P_{i}^{D} $. For a prosumer i, it is possible to write both quantities as a sum over the distributed energy resources ($ DER_{i} $) and loads ($ load_{i} $) of i :

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar radiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix \ref{appendixB} and in \cite{Lydia2014}.

Note that a prosumer i is defined by his zone $ Z_{i} $ as well as the sets $ DER_{i} $ and $ load_{i} $. That is, a prosumer can be configured to represent anything from a single wind-turbine for instance ($ DER_{i} = \{ WT_{0} \} $ and $ load_{i} = \emptyset $) to a pure load ($ DER_{i} = \emptyset $ and $ load_{i} = \{ L_{0} \} $) through more complex combinations. In practice, we use random configurations for the agents. In the rest of the paper, we use french weather data \cite{Infoclimat} starting in January 2006 and ending in December 2012, with a sampling frequency of three hours, and generate N timeseries of extra-production over this date range.

@@

This is the best contract value that the coalition S can afford giving the stability policy $\phi$ of the grid operator. Figure \ref{fig:Gaussian} shows how $ P_{S}^{CRCT \star} $ evolves according to the reliability parameter $ \phi $. For illustration, the range of $ \phi $ values is shown from 0 to 1, but in practice, only small values of $ \phi $ really make sense : $ \phi = 1 $ for instance means that coalitions can announce absolutely anything since the probability of producing less than any contract value is necessarily less than one by trivial definition of a probability. As visible on figure \ref{fig:Gaussian}, coalitions with high expected productions but presenting a high unpredictability are penalized and can only afford small contracts. The grid operator also specifies a lower bound $ P^{MIN} $ on the contract values as not to overload the market with unrealistic small coalitions. We thus characterized a valid coalition as one satisfying the two conditions :

@@

\begin{equation}
gain(S) = \int_{t_{0}}^{t_{k}} \lambda P_{S}^{CRCT \star} dt = P_{S}^{CRCT \star} \int_{t_{0}}^{t_{k}} \lambda dt
\end{equation} 
(since $ P_{S}^{CRCT \star} $ is supposed to be a constant rate over the contracted period). Using $ gain(S) $ directly as a utility function suffers a major drawback. It is indeed not a concave function of the coalitions' sizes, meaning that coalitions can grow as large as the number of agents allows it, without any counterbalance effects. Such a model, that virtually allows infinitely large coalitions and contract values, is in practice not realistic. There are indeed costs (communication costs for instance) that increase with the coalitions sizes. We take this observation into account by rescaling the utility of a coalition S by its size in term of number of agents ($|S|$):

\begin{equation}
\mathcal{U}(S) = \left\{ \begin{array}{lll}
							\dfrac{1}{|S|^{\alpha}} \dfrac{ P_{S}^{CRCT \star} }{P^{MAX}},\ if\ S\ is\ valid, \\
							0,\ if\ S\ is\ not\ valid
						 \end{array}
				  \right.
\end{equation}
where parameter $ \alpha $ controls to what extent the size of a coalition impacts its utility, and $ P^{MAX} $ is a normalizing factor. $ P^{MAX} $ can be seen as the maximum production which can be injected in the grid. Based on $ \mathcal{U} $, the marginal contribution of an agent i can be expressed as $ \delta_{S}(i) = \mathcal{U}(S+\{i\}) - \mathcal{U}(S) $. A coalition S has thus an interest in adding an additional agent i if this marginal contribution is positive :

@@

If $ \alpha $ is set to zero, agents are added as long as they increase the contract value of the coalition. If $ \alpha $ is greater than zero, additional agents have to increase the contract value by some factor. We relate now $ \alpha $ to the mean sizes of the coalitions $ \bar{N} $ by solving the following equation :

@@

Figure \ref{fig:mean_approx} shows how $ \alpha^{\star} $ and the utility function evolves according to the mean size of the coalitions $ \bar{N} $. These curves are only valid in the simplified Gaussian example considered here. Nonetheless, they will provide some guidance when using real data. As can be pointed out, the purpose of $ \mathcal{U} $ is not to study coalitions stability against player defection which could be done through game theory, nor to redistribute the coalition's utility in terms of individual payoffs. But we aim to design $ \mathcal{U} $ as a measure of how good a given coalition is according to our criteria.

@@

@@

\subsection{Representing the correlation structure}
As seen in section \ref{sec:notations}, the variance of the aggregated production impacts directly the contract values, and depends on the covariances between the agents productions. We argue here that, by having some representation of the correlation structure between the agents, the search landscape for high utility coalitions could be reduced, such that good coalitions are more likely to be found quickly. Usually, this correlation structure is formalized with a covariance matrix or a correlation matrix that contains all the correlation coefficients between the agents : $ M = (\rho_{ij})_{\forall i,j \in \mathcal{A}^{2}}$. By using a metric to map this matrix in a weighted adjacency matrix (see section \ref{sec:related}), it is possible to obtain a graph representation of the correlation relationships between the agents. In the following, we use two opposite distance metrics for this mapping :

@@

Clearly, $ d^{1} $ (resp. $ d^{2} $) maps two correlated series as close points (resp. distant) while two uncorrelated series are distant (resp. close). These metrics enable us to compute a correlation graph $ G_{1} = (\mathcal{A}, E_{1}) $ and a "de-correlation" graph $ G_{2} = (\mathcal{A}, E_{2} ) $. For any i and j, the weight of the edge $ e_{ij} $ is $ d_{ij}^{1} $ in $ G_{1} $ and $ d_{ij}^{2} $ in $ G_{2} $. In both cases, we want to keep only the edges which weights are located in the lower tail of the distance distributions. In other words, we want to compute the $ \epsilon$-graphs of $ G_{1} $ and $ G_{2} $ such that only meaningful edges remain. Selecting the filter $ \epsilon $ is an important point affecting the landscape search for the coalition formation. Unfortunately, there seems to be no clear consensus in the literature on how to select such a threshold. We will see later in this section that cliques in $ G_{2} $ are potential seeds for the coalitions. Since we want to generate $ N_{COAL} $ coalitions, we need at least $ N_{COAL} $ cliques of a given size to start. Besides, since we consider coalitions as disjoint, the starting cliques should be non overlapping. We select our optimal threshold for $ G_{2} $ as :

@@

Nevertheless de-correlation seems like a more complex concept than correlation in the sense that there is not even a partial notion of transitivity when it comes to it. Therefore, the clustering coefficients of $ G_{1}^{\epsilon} $ is much higher than the one of $ G_{2}^{\epsilon} $. This can be seen as another formulation of \cite{Garas2008} on the structural roles of weak and strong links on financial correlation graphs. Strong links, accounting for strong correlation relationships, are responsible for the clustering, while weak links provide the connectivity between clusters. Searching for clusters in $ G_{2}^{\epsilon} $ and hoping that this strategy will provide a nice coalition structure of internally uncorrelated coalitions seems thus pointless. Consider now a clique in  $ G_{2}^{\epsilon} $, which is a complete subgraph of $ G_{2}^{\epsilon} $. Since there is a link for every pairs of nodes, we know, by construction, that a clique has a mean correlation and a maximum correlation less than $ \epsilon $.

\begin{figure}
\includegraphics[scale=.5]{cliques_histo}
\caption{Histograms of utility values for coalitions of size 3. Red bars stands for cliques in the decorrelation graph, and blue bars represents all the other possible triplets. As visible, cliques tend to exhibit higher utilities than randomly selected coalitions.}
\label{fig:histo_cliques}
\end{figure}

Figure \ref{fig:histo_cliques} shows the distributions of the utility values for cliques of size 3 (triangles) in $ G_{2}^{\epsilon^{\star}} $ and for all the other possible triplets of agents. It is clearly visible that cliques tend to exhibit higher utilities because of their de-correlation property. Choosing cliques in $ G_{2}^{\epsilon^{\star}} $ as coalitions seems therefore appealing. Nevertheless, the quality of the results seems to decrease as the sizes of the cliques increase. Indeed, the larger the desired cliques, the more dense $ G_{2}^{\epsilon^{\star}} $ becomes (see equation \ref{epsilon_star}). There is a point where cliques results more from noisy edges than true de-correlation, which decreases the quality of the results. Directly mapping cliques to coalitions by this de-correlation oriented approach is thus not sufficient. It is indeed possible that adding agents to these cliques has the combined effect of increasing the expected production while decreasing its stability. The question revolves around measuring the benefits of this production surplus compared to the disadvantage of having coalition with high volatility. This can be quantified by the marginal benefit in equation \ref{eq:marginal_benefit}.

@@

\begin{itemize}
\item \textbf{Random sampling of coalitions :} Coalitions are formed randomly without any other constraint that the desired size. This enables us to have an idea about the distributions of utility values for coalitions of a given size.
\item \textbf{Random sampling of coalition structures :} Coalition structures are sampled randomly by shuffling and random divisions of the agents. Algorithm \ref{alg:algo2} uses such a sampling and returns the highest utility coalition structure sampled. We refer to his algorithm as "\textit{random}" in the following.
\item \textbf{Correlated :} Opposite version of our algorithm. It performs a community detection on the correlation graph $ G_{1} $. The resulting coalitions have thus very high internal correlations. We thus expect this algorithm to perform very bad compared to the others. See algorithm \ref{alg:algo3}.
\end{itemize}

@@

\begin{figure}
\includegraphics[scale=.5]{real_utility_2}
\caption{Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to equation \ref{eq:alpha_star} in order to favor 40 agents coalitions.}
\label{fig:real_utility2}
\end{figure}

Figure \ref{fig:coalitions} shows the coalitions formed with the considered algorithms in the contract value / volatility space. The color map in the background indicates regions where we expect high utilities (red) and the ones where we expect very poor utility values (blue). The bottom right corner, with high contract values and low volatilities, is therefore the region where we wish to form our coalitions. A single coalition is represented by a marker and the color and shape of a marker indicates by which algorithm the coalition has been formed. Besides, the sizes of the coalitions are indicated on the markers, and the marker size is also proportional to the coalition size. We can see that the utility function results in approximatively balanced coalitions. Small yellow markers indicates the gravity centers of their respective coalition structures. The coalitions of correlated agents (green squares) are clearly of poor quality according to our criteria since they can only afford small production contracts, and with a very high volatility. The decorrelated coalitions (blue dots) are closer to the bottom right corner indicating a much better quality in term of productivity over volatility ratio. The black dotted line indicates the mean values for the random coalitions sampling technique. Each small dot stands for the mean position of all sampled coalitions of this given size. Variances are not indicated for readability, but are usually quite large since this sampling only takes the size as a constraint. We can see that as coalitions get larger, they tend to increase on average their contract values, but at the price of a higher volatility. The results of the random coalition structure sampling are shown with the red ellipses that represent the distribution of the gravity centers of the sampled structures. Since the center of the ellipses stands for the mean and each ellipse adds one standard deviation, more than $ 99\% $ of the sampled gravity centers are within the largest ellipse. The small yellow dot below the ellipses indicates the gravity center of our solution. It is thus visible that our greedy graph based algorithm is able to find a quite good coalition structure in terms of volatility and contract values.

\begin{figure}
\includegraphics[scale=.5]{./figures/coalitions/coalitions6}
\caption{Coalitions formed in the (contract value, volatility) space. The color map indicates qualities of portions of the plane. The closer to red the better (high contract values with small volatility). On the opposite, blue areas show poor quality (small contract values with high volatility). Blue dots stand for the decorrelated coalitions that we formed while green squares show correlated coalitions. The smaller yellow markers stand for the gravity centers of the coalition structures. The black dotted line shows how contract values and volatility evolve  when the size of the coalitions increases (a subset of the points are labeled by the size of the coalition they represent). Each point is the average over $ 10^{5} $ unconstrained draws of a random coalition. As we can also draw random coalition structures, we show the distribution of their gravity centers by the red ellipses (center is the mean, and each ellipse corresponds to one  standard deviation).}
\label{fig:coalitions}
\end{figure}
